---
title: 'Coursera - PML: Course Project'
author: "André Marinho"
date: "14/08/2020"
output: 
        html_document:
                keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0. Project Goal

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

## 1. Load Libraries

```{r message=FALSE}
library(randomForest)
library(rattle)
library(rpart.plot)
library(rpart)
library(ggplot2)
library(lattice)
library(caret)
library(gbm)
```

## 2. Load and Clean Data

```{r}
set.seed(14082020) # reproducibility

data_training <- read.csv(file="dataset/pml-training.csv", na.strings=c("NA", "#DIV/0!", ""), header=TRUE)

data_testing <- read.csv(file="dataset/pml-testing.csv", na.strings=c("NA", "#DIV/0!", ""), header=TRUE)

# Remove missing values

data_training <- data_training[, colSums(is.na(data_training)) == 0]

data_testing <- data_testing[, colSums(is.na(data_testing)) == 0]
```

## 3. Resample Datasets

```{r}
# Training data resampling: remove unnecessary columns (1:7)

data_training <- data_training[, -c(1:7)]

# Testing data resampling: remove unnecessary columns (1:7), and the "problem_id" column (of the Coursera assignment project)

data_testing <- data_testing[, -c(1:7)]
data_testing <- data_testing[, -c(length(data_testing))]
```

### 3.1. Create a validation sample

```{r}
set.seed(14082020) # reproducibility

# 70% training; 30% validation

training <- createDataPartition(data_training$classe, p=0.7, list=FALSE)
data_training <- data_training[training, ]
data_validation <- data_training[-training, ]

dim(data_training)
dim(data_validation)
dim(data_testing)
```

### 3.2. Analyse Correlated Features

```{r}
corr <- cor(data_training[, -length(data_training)])

# Highly correlated features, cutoff = 0.8

corr_high <- findCorrelation(corr, cutoff=0.8)
names(data_training)[corr_high]
```

## 4. Build Model 1: Decision Tree

```{r}
set.seed(14082020) # reproducibility
decision_tree <- rpart(classe ~., data=data_training, method="class")

# Plot the tree
rpart.plot(decision_tree)
```

## 5. Model 1 Results

```{r}
decision_tree_pred <- predict(decision_tree, data_validation, type="class")

confusion_matrix <- confusionMatrix(decision_tree_pred, data_validation$classe)
confusion_matrix

plot(confusion_matrix$table, col=confusion_matrix$byClass,
     main=paste("Decision Tree Accuracy = ",
                round(confusion_matrix$overall['Accuracy'], 4)))
```

## 5. Build Model 2: Random Forests

```{r cache=TRUE}
set.seed(14082020) # reproducibility
random_forests <- randomForest(classe ~., data=data_training)

# Plot the random forests
plot(random_forests)
```

## 6. Model 2 Results

```{r cache=TRUE}
random_forests_pred <- predict(random_forests, data_validation, type="class")

cm_rf <- confusionMatrix(random_forests_pred, data_validation$classe)
cm_rf

plot(cm_rf$table, col=cm_rf$byClass, 
     main=paste("Random Forests Accuracy = ",
                round(cm_rf$overall['Accuracy'], 4)))
```

## 7. Build Model 3: Generalized Boosted Regression

```{r}
set.seed(14082020) # reproducibility
library(caret)
control_boost <- trainControl(method="repeatedcv",
                              number=5,
                              repeats=1)
gbr <- train(classe ~., data=data_training, method="gbm",
             trControl=control_boost, verbose=FALSE)
gbr$finalModel
```

## 8. Model 3 Results

```{r cache=TRUE}
gbm_pred <- predict(gbr, newdata=data_validation)

cm_gbm <- confusionMatrix(gbm_pred, data_validation$classe)
cm_gbm

#plot(gbm_pred, ylim=c(0.9, 1))
```

## 9. Classify Data

Best model based on accuracy -> Random Forests. Then, use random_forests to predict the test data values.

```{r cache=TRUE}
data_predict <- predict(random_forests, newdata=data_testing)
data_predict
```